Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-01 paper/paper.qmd
Improve decision thresholds / 95th risk pct

prevalance of cdi. prevalance of severe cdi outcomes.

Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.

prediction models based on EHR for whether infection occurs in the first place already in use.
so how about predicting severity of infections to guide treatment?
OTUs vs EHRs to predict severity.

@@

We first set out to train the best models possible for each severity definition.
Not all samples have labels available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review (@tbl-counts).

However, comparisons across these definitions is not fair when using different
subsets of the data for each definition.
To better compare the model performances across different severity definitions,
we selected the intersection of samples (`r n_cases_int`) that had labels for
all four severity definitions and repeated the model training and evaluation
process.

```{r sample_counts}
#| label: tbl-counts
#| tbl-pos: h
#| tbl-cap: "Dataset sample counts"
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
#| layout-ncol: 2
format_table <- function(tab) {
  tab %>%
    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
          format.args = list(big.mark = ',')) %>% 
    kable_styling(font_size = table_font_size)
}
read_csv(here('results','count_table_full.csv')) %>% 
  format_table()
read_csv(here('results','count_table_int.csv')) %>% 
  format_table()
```

@@

## Estimating clinical benefit

it's not enough for models to perform well to deploy it in a clinical setting;
benefit over current practices must be shown.
table of NNS at 95th pct of risk for each outcome x dataset (@tbl-risk).

@@

NNT for Fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is Vancomycin
because it's cheaper than Fidaxomicin, even though IDSA recommends Fidaxomicin.

pragmatic vs attributable severity. table?

@@

```{r risk_95th_pct_decision_thresholds, message=FALSE}
#| label: tbl-risk
#| tbl-pos: h
#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
#|   The confusion matrix was computed at the 95th percentile of risk for each
#|    severity prediction model, which corresponds to 5% of cases predicted to 
#|    be severe. The number needed to screen (NNS) to identify one true positive
#|    is the reciprocal of precision.
#|    "
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
risk_thresh %>% 
  filter(Dataset == 'Full') %>% 
  select(-Dataset) %>% 
  kable() %>% 
  kable_styling(font_size = table_font_size)
risk_thresh %>% 
  filter(Dataset == 'Intersection') %>% 
  select(-Dataset) %>% 
  kable() %>% 
  kable_styling(font_size = table_font_size)
```

@@

@@

Discuss important OTUs. which ones concord with literature, which ones may be new.

Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, we could make a 
case to routinely sequence and profile the microbial communities of CDI patients.

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was resuspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

@@

Samples stored in DNA genotek buffer were thawed from the -80&deg;C, vortexed,
and then transferred to a 96-well bead beating plate for DNA extractions.
DNA was extracted using the DNeasy
Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system
(Eppendorf).
The V4 region of the 16S rRNA gene was amplified with the AccuPrime
Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as
previously described [@kozich_development_2013].
Each library preparation plate for
sequencing contained a negative control (water) and mock community control
(ZymoBIOMICS microbial community DNA standards).
The PCR amplicons were
normalized (SequalPrep normalizatin plate kit from Thermo Fisher Scientific),
pooled and quantified (KAPA library quantification kit from KAPA Biosystems),
and sequenced with the MiSeq system (Illumina).

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in 
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
<!-- TODO supplementary figure with alpha and beta diversity & significance -->

@@

## Model training and evaluation

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation importance was then performed to determine which OTUs contributed
most to model performance.
We reported OTUs with a significant permutation test in at least 75 of the 100
models.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left panels, models were trained on the full dataset, with different
numbers of samples available for each severity definition.
In the right panels, models were trained on the intersection of samples with all
labels available for each definition.
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus identical values are expected.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets,
and the area under the balanced precision-recall curve (AUBPRC) for the test sets.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
Each point is the median with tails as the 95% CI.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
OTU 120 (*Pseudomonas*) is not shown for IDSA severity on the full dataset nor
on the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). The vertical dashed line 
represents the limit of detection.
"}
![](figures/feature-importance.png)
:::

>>>>

prevalence of cdi. prevalence of severe cdi outcomes.

Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.
Contribution of the gut microbiome.

prediction models based on EHR for whether infection occurs in the first place already in use.
so how about predicting severity of infections to guide treatment?
also models on EHR to predict adverse outcomes (Li, Rao).
OTUs vs EHRs to predict severity.
CDI severity prediction models could be deployed to screen patients at risk and
guide clinicians to consider prescribing a different course of treatment.
When paired with treatment options that may reduce risk of severity, deploying
prediction models can guide clinician decision-making to improve patient outcomes
while minimizing unnecessary harms.

@@

We first set out to train the best models possible for each severity definition.
Not all samples have labels available for all four severity definitions due to
missing data for some patient lab values and incomplete chart review (@fig-flowchart B),
thus each severity definition has a different number of samples (@tbl-counts)
when using as many samples as possible. We refer to these as the full datasets.
Random forest models were trained on repeated 100 splits of the datasets into
training and test sets, and performance was evaluated on the held-out
test set using the area under the receiver-operator characteristic curve (AUROC)
the area under the balanced precision-recall curve (AUBPRC).

However, comparisons across these definitions is not fair when using different
subsets of the data for each definition.
To better compare the model performances across different severity definitions,
we selected the intersection of samples (n=`r n_cases_int`) that had labels for
all four severity definitions and repeated the model training and evaluation
process.

```{r sample_counts}
#| label: tbl-counts
#| tbl-pos: h
#| tbl-cap: "**Sample counts and proportion of severe cases.** Each severity 
#|   definition has a different number of patient samples available, as well as
#|   a different proportion of cases labelled as severe.
#| "
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
#| layout-ncol: 2
format_table <- function(tab) {
  tab %>%
    kable(col.names = c(' ', 'IDSA', 'Attributable', 'All-cause', "Pragmatic"),
          format.args = list(big.mark = ',')) %>% 
    kable_styling(font_size = table_font_size)
}
read_csv(here('results','count_table_full.csv')) %>% 
  format_table()
read_csv(here('results','count_table_int.csv')) %>% 
  format_table()
```

@@

## Estimating clinical value

When considering the suitability of a model for deployment in clinical settings, 
the number needed to screen (NNS) is a highly relevant metric representing how
many patients must be predicted as severe by the model to identify one true positive.
We determined the decision threshold at the 95th percentile
of risk for each prediction model and computed the confusion matrix and several
performance metrics at this threshold including the NNS (@tbl-risk).
Among the models predicting severe outcomes, the models trained on the full
datasets performed best with an NNS of `r allcause_nns` for the all-cause
definition, `r attrib_nns` for the attributable definition, and 
`r pragmatic_nns` for the pragmatic definition.
For context, prior studies predicted CDI-attributable severity using whole
Electronic Health Record data and from a smaller set of 9 clinician-curated
variables, achieving precision values of 0.417 (NNS = `r 1/0.417`) for the EHR
model and 0.167 (NNS = `r 1/0.167`) for the curated model [@li_using_2019;
@rao_clostridium_2015].
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than prior curated models and it is the most clinically
relevant as physician chart review ensures that positively-labelled severe
outcomes are due to the CDI rather than other causes.

Even if a model performs well, it may not be useful in a clinical setting unless
it can guide clinicians to choose between treatment options.
At this time, we are not aware of any direct evidence that a particular
treatment reduces the risk of severe CDI outcomes.
However, with some assumptions we offer a proof-of-concept analysis of the
potential clinical value of OTU-based severity prediction models.
Paired with the NNS of these models,

Current clinical guidelines specify vancomycin and fidaxomicin as the standard
antibiotics to treat CDI, with a preference for fidaxomicin due to its higher
sustained resolution of CDI and reduced risk of recurrence
[@stuart_clinical_2021].
However, fidaxomicin is considerably more expensive than vancomycin, resulting
in most CDI patients in the US being treated with vancomycin (TODO citation).
If fidaxomicin were shown to reduce the risk of severe CDI outcomes, it could be
preferentially prescribed to patients predicted to be at risk.
While we are not aware of evidence demonstrating the superiority of fidaxomicin
for reducing severe CDI outcomes,

@@

NNT for fidaxomicin, FMT, and/or Bezlotoxumab. Current standard is vancomycin
because it's cheaper than fidaxomicin, even though IDSA recommends fidaxomicin.

@@

```{r risk_95th_pct_decision_thresholds, message=FALSE}
#| label: tbl-risk
#| tbl-pos: h
#| tbl-cap: "**Predictive model performance at 95th percentile of risk.** 
#|   The confusion matrix was computed for the decision threshold at
#|   the 95th percentile of risk for each severity prediction model, 
#|   which corresponds to 5% of cases predicted to have a severe outcome. 
#|   The number needed to screen (NNS) to identify one true positive
#|   is the reciprocal of precision.
#|   "
#| tbl-subcap: 
#|   - "Full datasets"
#|   - "Intersection of samples with all labels available"
risk_thresh <- read_csv(here('results','decision_thresholds.csv'))
risk_thresh %>% 
  filter(Dataset == 'Full') %>% 
  select(-Dataset) %>% 
  kable() %>% 
  kable_styling(font_size = table_font_size)
risk_thresh %>% 
  filter(Dataset == 'Intersection') %>% 
  select(-Dataset) %>% 
  kable() %>% 
  kable_styling(font_size = table_font_size)
```

@@

Performance

Discuss important OTUs. which ones concord with literature, which ones may be new.
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.

@@

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @stuart_clinical_2021.

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

@@

This study was approved by the University of Michigan Institutional Review
Board. All patient samples were collected by the University of Michigan Health
System from January 2016 through December 2017.
Stool samples that had unformed stool consistency were tested for *C. difficile* by
the clinical microbiology lab with a two-step algorithm that included detection
of *C. difficile* glutamate dehydrogenase and toxins A and B by enzyme
immunoassay with reflex to PCR for the *tcdB* gene when results were discordant.
1,517 stool samples were collected from patients diagnosed with a CDI.
Leftover stool samples that were sent to the clinical microbiology lab were
collected and split into different aliquots.
For 16S sequencing, the aliquot of stool was re-suspended in
DNA genotek stabilization buffer and then stored in the -80&deg;C freezer.
Only the first CDI sample per patient was used for subsequent ML analyses such
that no patient is represented more than once, resulting in a dataset of
`r n_cases_first` samples.

@@

Samples stored in DNA genotek buffer were thawed from the -80&deg;C, vortexed,
and then transferred to a 96-well bead beating plate for DNA extractions.
DNA was extracted using the DNeasy
Powersoil HTP 96 kit (Qiagen) and an EpMotion 5075 automated pipetting system
(Eppendorf).
The V4 region of the 16S rRNA gene was amplified with the AccuPrime
Pfx DNA polymerase (Thermo Fisher Scientific) using custom barcoded primers, as
previously described [@kozich_development_2013].
Each library preparation plate for
sequencing contained a negative control (water) and mock community control
(ZymoBIOMICS microbial community DNA standards).
The PCR amplicons were
normalized (SequalPrep normalization plate kit from Thermo Fisher Scientific),
pooled and quantified (KAPA library quantification kit from KAPA Biosystems),
and sequenced with the MiSeq system (Illumina).

All sequences were processed with mothur (v1.46) using the MiSeq SOP protocol
[@schloss_introducing_2009; @kozich_development_2013].
Paired sequencing reads were combined and
aligned with the SILVA (v132) reference database [@quast_silva_2013] and taxonomy was
assigned with a modified version of the Ribosomal Database Project reference
sequences (v16) [@cole_ribosomal_2014].
Sequences were clustered into _de novo_ OTUs with the OptiClust algorithm in
mothur [@westcott_opticlust_2017], resulting in `r num_otus` OTUs.
<!-- TODO supplementary figure with alpha and beta diversity & significance.
Samples were rarefied to 5,000 sequences per sample, repeated 1,000
times for alpha and beta diversity analysis.
 -->

@@

## Model training

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have full datasets which use all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were pre-processed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. 
During pre-processing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
No features had missing values and no features were perfectly correlated.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.

## Model evaluation

Model performance was calculated on the held-out test sets using the area under
the receiver-operator characteristic curve (AUROC) and the area under the
balanced precision-recall curve (AUBPRC).
Permutation feature importance was then performed to determine which OTUs
contributed most to model performance.
We reported OTUs with a significant permutation test in at least 75 of the 100
models.

@@

::: {#fig-performance fig-cap="**Performance of ML models.**
In the left panels, models were trained on the full dataset, with different
numbers of samples available for each severity definition.
In the right panels, models were trained on the intersection of samples with all
labels available for each definition.
Note that the intersection dataset for Attributable and Pragmatic severity have
exactly the same labels, thus identical values are expected.
**A)** 
Area under the receiver-operator characteristic curve (AUROC) for the test sets
and cross-validation folds of the training sets, and the area under the balanced
precision-recall curve (AUBPRC) for the test sets.
Each point is the median performance across 100 train/test splits with tails as
the 95% CI.
Nearly all pairs of definitions have significantly different performances on the
test set (P < 0.05) except for AUROC and AUBPRC of Attributable vs. Pragmatic on
the intersection dataset (as they are identical), AUROC of Attributable vs.
All-cause on the full dataset, and AUROC of Attributable vs. IDSA on the full
dataset.
**B)** Receiver-operator characteristic curves for the test sets.
Mean specificity is reported at each sensitivity value, 
with ribbons as the 95% CI.
**C)** Balanced precision-recall curves for the test sets.
Mean balanced precision is reported at each recall value, 
with ribbons as the 95% CI.
"}
![](figures/ml-performance.png)
:::

::: {#fig-features fig-cap="**Feature importance.**
**A)** Feature importance via permutation test.
For each OTU, the order of samples was randomized in the test set 100 times and
the performance was re-calculated to estimate the permutation performance.
An OTU was considered important if the performance decreased when the OTU was
permuted in at least 75% of the models.
OTUs with a greater difference in AUROC (actual performance minus permutation
performance) are more important.
Left: models were trained on the full datasets, with different numbers of
samples available for each severity definition.
Right: models were trained on the intersection of samples with all labels
available for each definition. Note that Attributable and Pragmatic severity are
exactly the same for the intersection dataset.
*Pseudomonas* (OTU 120) is not shown for IDSA severity in the full datasets nor
in the intersection dataset because it was removed during pre-processing due to
having near-zero variance.
**B)** Log~10~-transformed median relative abundances of the most important OTUs
on the full datasets, grouped by severity (shape). 
The vertical dashed line is the limit of detection.
"}
![](figures/feature-importance.png)
:::
